Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/04_Discussion.Rmd
another important point concerning the simulation

Concerning the question of cultural distance, the expectation based on Cultural transmission theory was a clear correlation of the "cultural" SED distances for both variables *burial type* and *burial construction* with spatial distance. The simulation, though with this implementation and parametrisation for sure rather qualitatively than quantitatively, visualises how a high degree of interaction should translate to a higher degree of cultural similarity across human populations (Figure \ref{simulation}). And as the European Bronze Age is usually depicted as a period of trade, travel and exchange [e.g. @kristiansenRiseBronzeAge2005; @sherrattWhatWouldBronzeAge1993] there was no reason to expect something else than Tobler's famous First Law of Geography: "everything is related to everything else, but near things are more related than distant things." [@toblerComputerMovieSimulating1970].

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation can be assumed based on the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

@@

3. All observations of cultural distance in this paper are based on only the two variables *burial type* and *burial construction*. This is by far not enough for any overarching conclusions about the general state of society in the Bronze Age.

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and introduced more variability for a -- at that time -- less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus Culture context. With the rise of the Urnfield Culture this phenomenon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development in early European history and was possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sun barks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if correct, only be considered on a big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the decision which burial tradition to adopt [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].

>>>>

Concerning the question of cultural distance, the expectation based on Cultural transmission theory was a clear correlation of the "cultural" SED distances for both variables *burial type* and *burial construction* with spatial distance. The simulation, though with this implementation and parametrisation for sure rather qualitatively than quantitatively, shows how an increase in the degree of interaction should translate to a higher degree of cultural similarity across human populations (Figure \ref{simulation}). And as the European Bronze Age is usually depicted as a period of trade, travel and exchange [e.g. @kristiansenRiseBronzeAge2005; @sherrattWhatWouldBronzeAge1993] there was no reason to expect something else than Tobler's famous First Law of Geography: "everything is related to everything else, but near things are more related than distant things." [@toblerComputerMovieSimulating1970].

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation is visible in the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

@@

3. The toy simulation based on @neimanStylisticVariationEvolutionary1995 suggests that even under the assumption of strong transmission between directly neighbouring groups, random and emergent effects can still cause 	fluctuation and temporary lack of correlation in human populations.

4. All observations of cultural distance in this paper are based on only the two variables *burial type* and *burial construction*. This is by far not enough for any overarching conclusions about the general state of society in the Bronze Age.

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and introduced more variability for a -- at that time -- less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus culture context. With the rise of the Urnfield culture this phenomenon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development and possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sun barks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if correct, only be considered on a big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the decision which burial tradition to adopt [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].
